\begin{abstract}

The rapid expansion of distributed systems and the growing sophistication of cyberattacks, particularly DDoS attacks, have created an urgent need for scalable and privacy-preserving security solutions. Traditional centralized intrusion detection systems face limitations in data privacy, communication overhead, and adaptability within heterogeneous environments. To address these challenges, this project explores the application of federated learning for anomaly-based detection of DDoS and network intrusions, with a focus on evaluating and improving the performance of various federated optimization algorithms. The study employs two datasets, CIC-IDS2017 and CIC-DDoS2019, which cover a broad range of DDoS and intrusion types. Multiple experimental setups were implemented with varying degrees of client heterogeneity and client counts to simulate realistic deployment scenarios. Models were trained using FedAvg, FedProx, FedAdam, and FedYogi, alongside autoencoder-based architectures for anomaly detection. Performance was assessed using key metrics such as accuracy and F1-score to capture predictive effectiveness. The experimental results demonstrate that federated learning is highly effective for DDoS detection, achieving comparative performance across testing setups. Among the tested algorithms, FedProx consistently outperformed FedAvg across heterogeneous conditions, while FedAdam showed notable improvements in large-scale, non-IID environments but underperformed in smaller client scenarios. A hybrid approach combining FedProx and FedAdam also achieved exceptional results, with peak results of 97.29\% accuracy and 97.29\% F1-score, which are comparable to centralized training even under highly non-IID conditions. Overall, this research provides empirical evidence that federated optimization techniques can enable scalable and privacy-preserving DDoS detection while highlighting the importance of aligning algorithm selection with the scale and heterogeneity of the deployment environment.

\end{abstract}
